---
title: "An Executable SpatialOmicsMSI Workflow"
author: "Spatial Omics Team"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{An Executable SpatialOmicsMSI Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
set.seed(20260811)
```

# Scope

This is the package's **single executable vignette**. Every table and figure
below is generated when the vignette is built; screenshots and pre-rendered
analysis figures are not used.

The compact fixture contains 2,500 measured kidney MSI pixels and 24 real m/z
features reconstructed from the public METASPACE acquisition
`2023-12-19_13h02m39s` associated with Kasarla et al. (2025). A fixed measured
field and a fixed variance-based feature subset are used only to keep package
installation and vignette execution fast. The generation script is
`scripts/prepare_vignette_fixture.R`, and provenance is recorded in
`inst/extdata/kasarla2025/SOURCE.md`.

This example supports a complete MSI computational workflow and a separate
ROI-level MALDI--LMD-LC-MS/MS validation. It does **not** contain matched
Visium expression, a registration-ready H&E image, or biological replicates;
therefore transcriptomic deconvolution, expression ligand--receptor analysis,
histology registration, and population-level inference are not claimed here.

```{r load-package}
library(SpatialOmicsMSI)

fixture <- system.file(
  "extdata", "kasarla2025", "kasarla_kidney_vignette_pixels.csv.gz",
  package = "SpatialOmicsMSI"
)
stopifnot(nzchar(fixture), file.exists(fixture))
```

# The five-step processing workflow

## 1. Validate and import pixel-level MSI

The input contract is one row per measured pixel, finite `x` and `y`
coordinates, and numeric columns named `mz_<mass>`. `pixel_id` provides a stable
join key. The pipeline validates this contract before analysis.

```{r import}
raw_pixels <- import_peakpicked_msi_csv(fixture)
c(
  measured_pixels = nrow(raw_pixels),
  mz_features = length(feature_columns(raw_pixels)),
  duplicated_coordinates = sum(duplicated(raw_pixels[c("x", "y")]))
)
```

## 2. QC, TIC normalization, and one log transform

The production path performs per-pixel TIC normalization followed by exactly
one `log10(x + 1)` transform. It returns analysis values separately from
coordinates and metadata so that downstream joins remain explicit.

```{r preprocess}
pipeline <- run_spatial_metabolomics_pipeline(
  fixture,
  bad_pixel_filter = FALSE,
  tic_normalize = TRUE,
  do_log = TRUE,
  do_scale = FALSE
)
unlist(pipeline$qc_summary)

pixels <- merge(
  pipeline$coordinates,
  pipeline$pixel_feature_matrix,
  by = "pixel_id",
  sort = FALSE
)
pixels <- pixels[match(pipeline$pixel_feature_matrix$pixel_id, pixels$pixel_id), ]
stopifnot(identical(pixels$pixel_id, pipeline$pixel_feature_matrix$pixel_id))
```

The map below is calculated from a real ion image, without interpolation or
background filling.

```{r ion-map, fig.cap="Highest-variance real MSI feature in the compact Kasarla kidney field."}
ion_feature <- feature_columns(pixels)[1]
ion_values <- pixels[[ion_feature]]
palette <- hcl.colors(64, "Viridis")
colour_index <- cut(ion_values, 64, include.lowest = TRUE, labels = FALSE)
plot(
  pixels$x, pixels$y,
  col = palette[colour_index], pch = 15, asp = 1,
  xlab = "MSI x", ylab = "MSI y", main = ion_feature
)
```

## 3. Construct the spatial graph and screen features

Moran's I has two valid roles: it can screen spatially structured features
before clustering, and the same cached result can later check whether
differential hits are spatially coherent. Here a symmetric Queen graph and 99
two-sided permutations make the executable example fast. A final study should
increase the permutation count (for example, 499 or more) and report its
attainable p-value resolution.

```{r moran}
moran <- compute_spatially_variable_metabolites(
  pixels,
  n_perm = 99,
  alternative = "two.sided",
  neighbor_method = "queen",
  seed = 20260811
)
head(moran[c("feature", "morans_i", "p_value", "adj_p_value", "n_effective")], 8)
```

For illustration, the twelve highest-ranked features are retained for domain
detection. This choice is an analysis parameter, not an anatomical assertion.

```{r feature-screen}
screened_features <- head(moran$feature, 12)
domain_input <- pixels[c("pixel_id", "x", "y", screened_features)]
```

## 4. Detect metabolic domains and neighborhood-composition niches

Domains group pixels by their own molecular profiles. Niches instead group
pixels by the **local composition of surrounding domains**. They are related
but not interchangeable labels.

```{r domains}
domains <- cluster_pixels(
  domain_input,
  k = 4,
  pca_components = 8,
  nstart = 25,
  seed = 20260811
)
domain_pixels <- domains$matrix
domain_pixels$domain_label <- paste0("domain_", domain_pixels$cluster)
table(domain_pixels$domain_label)
```

```{r domain-map, fig.cap="Data-driven metabolic domains; these are not anatomical ground truth."}
domain_colours <- setNames(
  hcl.colors(length(unique(domain_pixels$domain_label)), "Dark 3"),
  sort(unique(domain_pixels$domain_label))
)
plot(
  domain_pixels$x, domain_pixels$y,
  col = domain_colours[domain_pixels$domain_label], pch = 15, asp = 1,
  xlab = "MSI x", ylab = "MSI y", main = "Metabolic domains"
)
legend("topright", legend = names(domain_colours), fill = domain_colours, bty = "n")
```

The radius is expressed in this fixture's grid-coordinate units. Hellinger
transformation is the default for composition vectors because it handles zeros
without a pseudocount. Pixels below `min_neighbors` would receive `NA` niche
labels and remain counted in the exclusion summary.

```{r niches}
niches <- detect_spatial_niches(
  domain_pixels,
  domain_column = "domain_label",
  neighborhood = "radius",
  radius = 2.01,
  include_self = TRUE,
  min_neighbors = 3,
  transform = "hellinger",
  k_niches = 3,
  domain_alignment = "joint",
  seed = 20260811
)
niches$exclusion_summary
table(niches$matrix$niche_label, useNA = "ifany")
```

## 5. Define ROIs, sample subregions, and create standard outputs

There are two ROI paths in the package:

1. data-driven ROI selection retains one or more reviewed domains;
2. coordinate ROI selection applies independently defined rectangles.

This example demonstrates the first path. We retain the two largest domains as
separate ROI groups, then aggregate local tiles. Selecting domains does not
refit the clustering.

```{r roi-sampling}
domain_order <- names(sort(table(domain_pixels$cluster), decreasing = TRUE))
roi_mapping <- data.frame(
  cluster = as.integer(domain_order[1:2]),
  roi_id = c("reviewed_domain_A", "reviewed_domain_B")
)
roi_pixels <- define_rois(
  domain_pixels,
  mode = "cluster",
  cluster_mapping = roi_mapping
)

subregions <- sample_subregions(
  roi_pixels,
  grid_size = 5,
  min_pixels = 8,
  grid_scope = "roi"
)
table(subregions$sample_matrix$roi_id)
```

The standardized result objects can be written directly without reshaping.
`make_metaboanalyst_data()` creates a sample-by-feature table with explicit
group labels; it belongs in MetaboAnalyst's statistical-analysis input, not in
the one-column peak-list uploader.

```{r standard-outputs}
metaboanalyst_table <- make_metaboanalyst_data(
  subregions$sample_matrix,
  group_column = "roi_id"
)
metaboanalyst_table[1:4, 1:6]
```

# Downstream analysis tasks

## Exploratory ROI differential analysis

Because this compact fixture contains one acquisition field and no independent
animals, the following p-values are exploratory tile-level summaries and must
not be presented as population inference. Real cohort analysis should supply
subject and section columns or use the package's subject-aware spatial model.

```{r differential}
differential <- suppressWarnings(differential_region_analysis(
  subregions$sample_matrix,
  group_column = "roi_id"
))
head(differential[c(
  "feature", "group_a", "group_b", "effect_size", "p_value", "fdr",
  "inference_unit", "pseudoreplication_warning"
)], 8)
```

The precomputed Moran table can now be reused in its second role: spatial
coherence checking of ROI-associated features. This is supporting evidence,
not an independent test, because both results use the same MSI field.

```{r differential-moran}
spatially_checked <- merge(
  differential,
  moran[c("feature", "morans_i", "adj_p_value")],
  by = "feature",
  suffixes = c("_roi", "_moran")
)
spatially_checked <- spatially_checked[order(spatially_checked$fdr), ]
head(spatially_checked[c("feature", "effect_size", "fdr", "morans_i", "adj_p_value")], 8)
```

## Study-matched MALDI--LMD-LC-MS/MS spatial validation

Kasarla et al. reported MALDI-MSI and LMD-LC-MS/MS region summaries for kidney
cortex, medulla, and renal pelvis. These installed tables validate the
ROI-concordance function independently of the compact pixel-field ROI labels
above. They are study-matched serial-section measurements, not interchangeable
pixel-level inputs.

```{r lcms-validation}
bundle_dir <- system.file("extdata", "kasarla2025", package = "SpatialOmicsMSI")
maldi_roi <- read.csv(
  file.path(bundle_dir, "kasarla_kidney_maldi_roi_matrix.csv"),
  check.names = FALSE
)
lmd_lcms_roi <- read.csv(
  file.path(bundle_dir, "kasarla_kidney_lmd_lcms_roi_matrix.csv"),
  check.names = FALSE
)
feature_mapping <- read.csv(
  file.path(bundle_dir, "kasarla_kidney_feature_mapping.csv"),
  check.names = FALSE
)

concordance <- compare_msi_lcm_quantification(
  msi_matrix = maldi_roi,
  lcm_matrix = lmd_lcms_roi,
  feature_mapping = feature_mapping,
  id_columns = c("roi_id", "section_id"),
  method = "spearman",
  min_pairs = 3
)
head(concordance[c(
  "msi_feature", "lcm_feature", "n_pairs", "correlation", "p_value"
)], 8)
```

With only three aggregate anatomical regions, correlations are coarse
descriptive concordance measures; they do not establish molecular identity or
replace replicate-aware quantification. The full source and licensing record is
shipped beside the tables.

# What this vignette does not claim

- Domains and niches are computational structures until independently interpreted.
- Tiles and pixels are not biological replicates.
- Accurate m/z alone does not confirm a lipid or metabolite identity.
- A MetaboAnalyst lipid/pathway upload requires reviewed lipid names or identifiers;
  the package must not invent them from m/z alone.
- Cell-type and expression ligand--receptor functions require suitable matched
  expression or validated label inputs that are not present in this fixture.

The separate Shiny guide documents the graphical workflow and example-dataset
capabilities. It is intentionally not a second formal package vignette.

<details><summary>Session information</summary>

```{r session-info}
sessionInfo()
```

</details>
