---
title: "Spatial Metabolomics Workflow"
author: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Metabolomics Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## Biological question

This vignette outlines a reproducible workflow for analyzing spatial metabolomics data
to identify spatially variable metabolites and validate putative identifications using
orthogonal LC-MS/MS data.

## Example dataset (msiFlow mouse bladder)

We use a small example subset from the `ly6g_heterogeneity_signatures.zip` archive
hosted on Zenodo (https://zenodo.org/records/11913042) as a compact demonstration
dataset. For extended analysis, the full `region_group_analysis.zip` in the same
record can be used. Do not attempt to download the full ~3.1 GB archive in this
demo — the vignette provides instructions and code to retrieve and prepare data
but those chunks are marked `eval=FALSE`.

## Data provenance and LC-MS/MS accession

- MSI and histology image files: see `ly6g_heterogeneity_signatures.zip` on Zenodo (Zenodo record: 11913042). For a formal DOI pointing to related MSI/IFM resources, see https://doi.org/10.5281/zenodo.14184882.
- Extended region-level data: `region_group_analysis.zip` (available via the Zenodo record linked above) — this archive provides the larger dataset used for extended analyses; it is not the sole formal accession for all source data.
- LC-MS/MS dataset accession: MSV000096465. LC-MS/MS is referenced here as an orthogonal method for validating putative metabolite identities (matching by m/z, ppm and ion mode). We do not claim LC-MS/MS represents matched quantitative biological replicates for the MSI samples.

# Data Processing

The processing stage is split into five explicit steps. Each step documents the
expected input, example function calls (from this package), outputs, and QC/failure checks.

## Step 1: Input validation and MSI import

- Input: a peak-picked MSI CSV with pixel coordinates and `mz_` feature columns.
- Function calls:

```{r eval=FALSE}
# import from CSV
pixel_matrix <- import_peakpicked_msi_csv("path/to/peakpicked.csv")
required_columns(pixel_matrix, c("x", "y"), "MSI pixel matrix")
fcols <- feature_columns(pixel_matrix)
```

- Output: a `data.frame` with pixel metadata and numeric feature columns.
- QC/Failures: check for finite coordinates, at least a small number of features,
  and existence of declared columns.

## Step 2: Pixel/feature QC

- Input: `pixel_matrix` from Step 1.
- Function calls:

```{r eval=FALSE}
# filter bad pixels and low-information features
keep <- rowSums(as.matrix(pixel_matrix[fcols]) > 0, na.rm = TRUE) >= 1
filtered <- pixel_matrix[keep, , drop = FALSE]
# construct QC summary
qc_summary <- list(raw_pixels = nrow(pixel_matrix), retained_pixels = nrow(filtered))
```

- Output: filtered pixel matrix and `qc_summary` list.
- QC/Failures: ensure `retained_pixels > 0`; otherwise stop with informative message.

## Step 3: TIC normalization, log transformation and scaling

- Input: filtered pixel-feature matrix.
- Function calls:

```{r eval=FALSE}
normalized <- sweep(as.matrix(filtered[fcols]), 1, rowSums(filtered[fcols], na.rm=TRUE), "/")
normalized <- log10(normalized + 1)
normalized <- scale(normalized)
```

- Output: numeric matrix `normalized` ready for spatial analysis.
- QC/Failures: check for non-finite values after transforms and handle zeros/NA.

## Step 4: H&E/IF image registration and ROI definition

- Input: histology control points, MSI coordinates and ROI definitions.
- Function calls: (illustrative; heavy image processing steps disabled here)

```{r eval=FALSE}
# register histology coordinates to MSI coordinates
reg <- fit_histology_msi_registration(control_points)
coords <- transform_histology_coordinates(histology_coordinates, reg)
# define ROIs from clusters or coordinate ROI table
roi_matrix <- define_rois(normalized, mode = "coordinate", roi_table = roi_table)
```

- Output: spatial transforms, transformed coordinate table, and ROI labels mapped to pixels.
- QC/Failures: verify ROI overlap with pixel coverage; warn if ROIs contain few pixels.

## Step 5: Subregion aggregation and standardized output

- Input: normalized feature matrix and ROI definitions.
- Function calls:

```{r eval=FALSE}
clustered <- cluster_pixels(normalized, k = 3, pca_components = 10)
roi_matrix <- define_rois(clustered$matrix, mode = "cluster")
sampled <- sample_subregions(
  roi_matrix,
  grid_size = 5,
  min_pixels = 30
)
metabo_data <- make_metaboanalyst_data(sampled$sample_matrix)
write.csv(sampled$sample_matrix, "region_summary.csv", row.names = FALSE)
```

- Output: aggregated region-level summaries and a standardized sample matrix ready for downstream analysis.
- QC/Failures: check that aggregation did not produce empty regions; validate column names.

# Data Analysis

This section describes common analysis tasks performed after processing.

## Task 1: Spatially variable metabolites using Moran's I

- Input: pixel-feature matrix and coordinates.
- Function calls:

```{r eval=FALSE}
spatial <- compute_spatially_variable_metabolites(pixel_feature_matrix, coordinates)
```

- Output: a table of features with Moran's I and permutation p-values.
- QC/Failures: ensure permutation counts > 0; check p-value finite.

## Task 2: Spatial domains and ROI profiles

- Input: normalized matrix and optional image-derived features.
- Function calls:

```{r eval=FALSE}
clustered <- cluster_pixels(normalized, k = 3, pca_components = 10)
# sample_subregions can be used to summarize ROI profiles by spatial subregions
sampled <- sample_subregions(clustered$matrix, grid_size = 5, min_pixels = 30)
profiles <- sampled$sample_matrix
```

- Output: domain segmentation map and per-region sample profiles.

## Task 3: Differential analysis between tissue regions

- Input: region-aggregated sample matrix and group labels.
- Function calls:

```{r eval=FALSE}
metabo_data <- make_metaboanalyst_data(sampled$sample_matrix)
# export metabo_data to MetaboAnalyst or other differential analysis tools
# result <- run_differential_analysis_in_metaboanalyst(metabo_data)
# normalized <- normalize_metaboanalyst_result(result, source_name = "metaboanalyst")
```

- Output: a prepared data frame for differential analysis and normalized result import support.

## Task 4: MSI–LC-MS/MS cross-validation using m/z, ppm and ion mode

- Input: MSI feature table and LC-MS/MS identifications.
- Function calls:

```{r eval=FALSE}
matched <- cross_validate_msi_lcms(msi_features, lcms_features, ppm = 10)
```

- Output: matched table with ppm error and direction agreement.
- QC/Failures: apply numeric tolerance when comparing ppm boundaries; inspect unmatched features.

# Data downloads and example dataset

Quick example (small): download and extract `ly6g_heterogeneity_signatures.zip` from
Zenodo (https://zenodo.org/records/11913042). For extended analyses, obtain the
`region_group_analysis.zip` archive from the same record (large, ~3.1 GB) and
follow the package's `inst/` or `data/` ingestion functions to prepare data.

LC-MS/MS dataset accession used as example: MSV000096465

---

Session info and package citation (render-time):

```{r session, eval=FALSE}
sessionInfo()
citation("SpatialOmicsMSI")
```

<!-- end vignette -->
