---
title: "Reproducible Spatial Metabolomics: msiFlow Mouse Bladder"
author: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducible Spatial Metabolomics: msiFlow Mouse Bladder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette documents the real-data workflow used for the msiFlow mouse-bladder
dataset associated with Zenodo record
[14184882](https://doi.org/10.5281/zenodo.14184882) and LC-MS/MS accession
[MSV000096465](https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?accession=MSV000096465).
It separates MSI measurement, data-driven spatial characterization, accurate-mass
annotation, and fragment-level orthogonal support.

The MSI example consists of four infected-bladder samples, UPEC_12--UPEC_15.
The article describes four infected samples from female mice, so the subject-level
analysis uses `subject_id = sample_id` and `section_id = sample_id`. The exact animal
identifier is not embedded in the imzML metadata; all subject-level results therefore
remain exploratory at (n=4). Pixels and tiles are never treated as biological
replicates.

Important provenance limits are fixed before analysis:

- The supplied imzML files already contain 252 centroided, shared-axis MSI features.
  Loading them does **not** mean this package performed raw profile peak picking,
  alignment, or de-isotoping.
- Positive polarity is passed explicitly with
  `ion_mode_source = "inferred_from_annotation_adducts"`. It was not established
  from polarity CV metadata in these imzML files.
- Domains 0--2 are data-driven metabolic domains derived from the same MSI data.
  Comparing them is exploratory characterization and has circularity risk.
- Domain -1 means unclassified/background. It is retained for maps and QC but
  excluded from formal domain comparisons.
- This downloaded subset has no H&E image. The source study uses IFM on consecutive
  sections; this workflow must not be described as H&E validation.
- LC-MS/MS is a companion orthogonal dataset, not pixel-matched or demonstrably
  paired to the individual MSI section. Fragment evidence supports compatibility
  with a paper-reported assignment, not MSI--LC-MS quantitative correlation.

# Data Processing

The reproducible processing contract is strictly:

```text
Input -> 1. validation/provenance -> 2. imzML feature matrix
      -> 3. pixel/feature QC -> 4. TIC normalization + log1p
      -> 5. standardized output + manifest
```

## Step 1: input validation and provenance

For imzML input, supply the `.imzML` path, its same-basename `.ibd`, sample and
section identifiers, and an explicit ion mode. `annotation.tsv` is not used to
pre-filter MSI features; it is only used later for one-to-many putative
accurate-mass annotation.

```{r processing-step-1, eval=FALSE}
library(SpatialOmicsMSI)

root <- "data_raw/msiflow/ly6g_heterogeneity_signatures"
imzml <- file.path(root, "msi", "UPEC_12.imzML")
ibd <- sub("[.]imzML$", ".ibd", imzml)
stopifnot(file.exists(imzml), file.exists(ibd))

input_manifest <- make_pipeline_manifest(
  input_files = c(imzml, ibd),
  input_type = "shared_axis_centroided_imzML",
  ion_mode = "positive",
  ppm = 10,
  parameters = list(
    sample_id = "UPEC_12",
    section_id = "UPEC_12",
    ion_mode_source = "inferred_from_annotation_adducts"
  )
)
```

The provenance must state that polarity was not established from imzML CV metadata.

## Step 2: centroided imzML/ibd reading and feature-matrix construction

`load_centroided_msi_features()` verifies the companion `.ibd`, MS1 spectra, a
shared m/z axis, and intensity-array orientation. It retains full-precision m/z in
`feature_metadata`; stable unique column names do not replace scientific m/z values.

```{r processing-step-2, eval=FALSE}
upec12 <- load_centroided_msi_features(
  imzml_path = imzml,
  sample_id = "UPEC_12",
  section_id = "UPEC_12",
  ion_mode = "positive",
  ion_mode_source = "inferred_from_annotation_adducts"
)

pixel_feature_matrix <- upec12$pixel_feature_matrix
coordinates <- upec12$coordinates
feature_metadata <- upec12$feature_metadata
stopifnot(nrow(pixel_feature_matrix) == 7036L)
stopifnot(nrow(feature_metadata) == 252L)
```

The observed UPEC_12 output contains 7,036 pixels and all 252 measured shared-axis
features. UPEC_13, UPEC_14, and UPEC_15 also contain 252 features, and their numeric
m/z axes match UPEC_12 exactly (maximum observed axis difference: 0 ppm).

## Step 3: pixel and feature QC

QC is performed on measured features before annotation. Coordinates must be finite
and unique within a section; feature intensities must be finite and non-negative;
zero-TIC pixels cannot enter normalization.

```{r processing-step-3, eval=FALSE}
feature_columns <- feature_metadata$column_name
intensity <- as.matrix(pixel_feature_matrix[, feature_columns, drop = FALSE])

stopifnot(all(is.finite(coordinates$x)), all(is.finite(coordinates$y)))
stopifnot(!anyDuplicated(coordinates[c("section_id", "x", "y")]))
stopifnot(all(is.finite(intensity)), all(intensity >= 0))

tic <- rowSums(intensity)
keep_pixel <- is.finite(tic) & tic > 0
qc_summary <- data.frame(
  input_pixels = nrow(intensity),
  retained_pixels = sum(keep_pixel),
  removed_nonpositive_tic = sum(!keep_pixel),
  measured_features = ncol(intensity)
)
```

## Step 4: TIC normalization and log1p transformation

Each sample is normalized separately. Transformation order is recorded and fixed:
per-pixel TIC normalization first, then `log1p`. Samples are not joined into one
spatial graph.

```{r processing-step-4, eval=FALSE}
intensity <- intensity[keep_pixel, , drop = FALSE]
tic <- tic[keep_pixel]
tic_normalized <- intensity / tic
analysis_matrix <- log1p(tic_normalized)
stopifnot(all(is.finite(analysis_matrix)))

processing_parameters <- list(
  normalization = "per-pixel TIC",
  transformation = "log1p",
  transformation_order = "TIC normalization then log1p"
)
```

## Step 5: standardized outputs and manifest

One run writes the pixel-by-feature matrix, coordinates, feature metadata, QC,
parameters, and provenance. These outputs are sufficient to restart analysis without
re-reading raw binary arrays.

```{r processing-step-5, eval=FALSE}
standard_output <- "results/real_data/UPEC_12"
write_pipeline_outputs(
  output_dir = standard_output,
  pixel_matrix = pixel_feature_matrix[keep_pixel, , drop = FALSE],
  coordinates = coordinates[keep_pixel, , drop = FALSE],
  feature_metadata = feature_metadata,
  qc_summary = qc_summary,
  processing_parameters = processing_parameters,
  provenance = c(
    upec12$provenance,
    list(
      polarity_cv_metadata_established = FALSE,
      input_manifest = input_manifest
    )
  )
)
```

# Data Analysis

The workflow answers four prespecified questions. Annotation is downstream of MSI
measurement and never controls which of the 252 features enters spatial analysis.

## Task 1: UPEC_12 spatially variable metabolites

The primary graph is symmetric binary Queen adjacency at distance
(sqrt{2}); Rook adjacency at distance 1 is a sensitivity analysis. Isolated pixels
do not enter the Moran mean, denominator, or permutations.

```{r analysis-task-1, eval=FALSE}
queen <- build_spatial_neighbors(
  coordinates$x, coordinates$y,
  method = "queen", distance_threshold = sqrt(2),
  weights = "binary", symmetric = TRUE
)
rook <- build_spatial_neighbors(
  coordinates$x, coordinates$y,
  method = "rook", distance_threshold = 1,
  weights = "binary", symmetric = TRUE
)

queen_result <- compute_spatially_variable_metabolites(
  pixel_matrix = data.frame(coordinates[c("x", "y")], analysis_matrix),
  coordinates = coordinates,
  n_perm = 499,
  alternative = "two.sided",
  p_adjust_method = "BH",
  seed = 20260806,
  neighbors = queen
)
```

Observed diagnostics and results were:

| Metric | Queen | Rook |
|---|---:|---:|
| Total pixels | 7,036 | 7,036 |
| Effective pixels | 6,992 | 6,935 |
| Isolated pixels | 44 | 101 |
| Undirected edges | 19,291 | 10,223 |
| Connected components | 329 | 408 |
| BH-FDR significant features | 184 | 166 |

Queen/Rook Moran rankings had Spearman correlation 0.97285. Their significant sets
overlapped in 157 features with Jaccard 0.8135. With 499 permutations, the minimum
attainable permutation p-value is (1/(499+1)=0.002); values at 0.002 indicate the
permutation resolution limit, not zero probability.

## Task 2: data-driven metabolic domains

The archive's `umap_data.csv` labels are mapped by sample, x, and y with a strict
one-to-one coordinate check. Values 0--2 are called data-driven metabolic domains;
-1 remains unclassified/background.

```{r analysis-task-2, eval=FALSE}
domain_pixels <- map_spatial_domain_labels(
  pixel_matrix = pixel_feature_matrix,
  umap_path = file.path(root, "umap_data.csv"),
  sample_id = "UPEC_12"
)
stopifnot(!anyNA(domain_pixels$domain_id))

formal_domain_pixels <- domain_pixels$domain_id %in% 0:2
```

For UPEC_12, pixel counts were 485 in -1, 1,417 in domain 0, 688 in domain 1,
and 4,446 in domain 2. Because these domains were learned from the same MSI feature
space being characterized, domain contrasts have circularity risk and are not
independent anatomical ROI validation.

## Task 3: UPEC_12--UPEC_15 subject-paired domain characterization

Each subject-by-domain-by-feature value is first reduced to one mean. Domain 0 versus
1, 0 versus 2, and 1 versus 2 are then evaluated with subject-paired tests; pixels
and tiles never enter the replicate count.

```{r analysis-task-3, eval=FALSE}
# `subject_domain_matrix` has one row per subject x domain and feature columns.
paired_domains <- differential_region_analysis(
  sample_matrix = subject_domain_matrix,
  group_column = "domain_id",
  subject_column = "subject_id",
  section_column = "section_id",
  p_adjust_method = "BH"
)
```

All four samples had complete coordinate/domain mappings. At (n=4), the exploratory
BH-FDR counts were 59 features for domain 0 versus 1, 107 for domain 0 versus 2,
and 0 for domain 1 versus 2. These small-sample paired results characterize the
published data-driven domains; they do not establish anatomical regions or generalize
beyond this proof-of-concept cohort.

## Task 4: PC(O-16:0_20:4) fragment-level orthogonal support

Diagnostic ions are supplied explicitly. The function never assigns chemistry from
an unlabelled peak. A core scan must satisfy both selected-precursor tolerance and
isolation-window coverage. Profile-derived local maxima retain their extraction
method, parameters, and software version.

```{r analysis-task-4, eval=FALSE}
pos_ions <- data.frame(
  label = "phosphocholine_184_07",
  target_mz = 184.07,
  evidence_role = "positive_mode_PC_class_fragment"
)
pos_support <- validate_msms_fragment_evidence(
  precursor_scans = pos_scan_metadata,
  fragment_peaks = pos_fragment_peaks,
  diagnostic_ions = pos_ions,
  precursor_target = 768.5902,
  precursor_ppm_tolerance = 10,
  fragment_tolerance = 0.02,
  fragment_tolerance_unit = "Da",
  min_relative_intensity = 0.01,
  min_core_scans = 2,
  required_diagnostic_labels = "phosphocholine_184_07"
)

neg_ions <- data.frame(
  label = c("M_minus_CH3_752_56", "FA_20_4_303_23",
            "FA_20_4_minus_CO2_259_24", "dLPC_16_0_466_33"),
  target_mz = c(752.56, 303.23, 259.24, 466.33),
  evidence_role = c("additional_consistency", "20:4_acyl_chain",
                    "additional_20:4_consistency", "16:0_lyso_fragment")
)
neg_support <- validate_msms_fragment_evidence(
  precursor_scans = neg_scan_metadata,
  fragment_peaks = neg_fragment_peaks,
  diagnostic_ions = neg_ions,
  precursor_target = 812.5811,
  precursor_ppm_tolerance = 10,
  fragment_tolerance = 0.02,
  fragment_tolerance_unit = "Da",
  min_relative_intensity = 0.01,
  min_core_scans = 2,
  required_diagnostic_labels = c("FA_20_4_303_23", "dLPC_16_0_466_33")
)
```

The real POS file produced 17 selected-ion matches for 768.5902 and 10 core scans;
184.07 reached repeated diagnostic support in all 10 core scans. The NEG file
produced six selected-ion matches for the formate adduct 812.5811 and five core
scans. Core scans meeting the 1% base-peak threshold numbered five for 752.56,
four for 303.23, three for 259.24, and four for 466.33. Thus POS PC-class support,
NEG 20:4 support, NEG 16:0 support, and the explicit overall rule were all `TRUE`.

The recommended result label is `fragment_level_orthogonal_support`. The source
mzML marks these MS2 spectra as profile; local maxima were derived with
`scipy.signal.find_peaks(prominence=0)` and are not vendor centroid data. The combined
evidence is consistent with the paper-reported PC(O-16:0_20:4) assignment, but the
LC-MS/MS dataset is companion rather than pixel-level or same-sample paired evidence.

Putative accurate-mass annotations remain one-to-many and must retain the label
`putative_accurate_mass_annotation`. Neither accurate-mass matching nor fragment-level
orthogonal support should be converted into an unqualified structure claim.

```{r session-info}
sessionInfo()
```
